Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2020-05-07 manuscript/manuscript.tex
incorporated grahams comments

In our analysis of the \textcite{Barghi2019-qy} data, we describe our statistic
$G(t)$ as a lower bound for two reasons: (1) the population is sequenced every
ten generations, meaning the temporal covariances between adjacent generations
cannot contribute to the numerator of $G(t)$ but contribute to the denominator,
and (2) the estimate of $G(t)$ ignores linked selection's contribution to the
\gc{per-generation} variance in allele frequency change (this is the
difference between \gc{our} $G(t)$ and
$G'(t)$ \gc{estimators}, the latter of which includes these variance terms
\parencite{Buffalo2019-io}).  To verify that $G(t)$ estimated every ten
generations is indeed a lower bound, we used a simulation procedure similar to
the exponential fitness function simulations (described in Supplementary
Material Section \ref{supp:genetic-arch}), and calculated the temporal
covariances and $G(t)$ both each generation, and every ten generations. Unlike
the simulations described in \ref{supp:genetic-arch}, we began selection at
$10N$ generations, and used trait $V_A = 0.01$ and targeted $L = 1000$ sites
affecting the trait.

@@

We simulated background selection in a 50 megabase region, where deleterious
alleles are randomly introduced by mutation. Following background selection
literature
\parencite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1994-oh,Hudson1995-xc}, we
parameterize the mutation rate as the total number of deleterious mutations
introduced per diploid genome, per generation, and simulate values $U = \{0.5,
1.0, 1.5\}$. We also vary the strength of selection against the deleterious
mutations, $s = \{0.0, 0.01, 0.05, 0.1\}$ (where $s=0$ is a neutral control),
as well as different recombination rates ($r_{bp} = \{10^{-7}, 10^{-8}\}$).
Like other simulations, we burnin the population for $10N$ generations under
backgrounds selection. Overall, we find background selection does create
temporal covariance (Supplementary Material Figure
\ref{suppfig:supp-bgs-covs}), which are stronger under (1) higher deleterious
mutation rates and (2) larger selection coefficients. This latter point
initially seems at odds with background selection theory, as the level of
pairwise diversity in a region under \gc{stroong} background selection is invariant with
respect to the selection coefficient. However, looking at the background
selection $G(t)$ trajectories, we find that over time, the impact of linked
selection created by background selection appears to trend towards an
equilibrium in the $r_{BP} = 10^{-7}$ subfigure, and reaches an equilibrium in
the $r_{BP} = 10^{-8}$ subfigure \gc{that seems reasonably invariant
  to the choice of $s$} (Supplementary Material Figure
\ref{suppfig:supp-bgs-g}). \gc{We believe that these observations can
  be reconciled by $Cov(\Delta ap_t,~ \Delta p_{t^{\prime}})$ being larger for larger
  $s$ when $|t^{\prime}-t|$ is small, but also decaying more rapidly
  with $|t^{\prime}-t|$, such that the overall contribution of selection to allele frequency variance is invariant to $s$ (for strong background selection).} However, further future work is needed to fully
explore and understand the temporal covariance dynamics of background
selection.

@@

Additionally, we investigated whether background selection can create
convergence correlations between two replicate populations. Much like the
exponential directional selection and Gaussian stabilizing selection
simulations, we burned in a population for $10N$ generations with background
selection, which continued after the population was split into two replicate
populations. These simulations fixed $U = 1.0$, $r_{BP} = 10^{-8}$, and varied
the replicate population size $n = \{200, 1000\}$. We find that background
selection can create convergence correlations (Supplementary Material Figure
\ref{suppfig:supp-bgs-converg}). We find the convergence correlation is weaker
in smaller replicate population sizes, as there are fewer shared haplotypes
carrying \gc{the same} deleterious alleles between the two populations.

\gc{While investigating} whether different selection coefficients converged to
the same levels of $G(t)$, we extended the number of generations we calculated
our statistics on (e.g. in earlier exponential and Gaussian stabilizing
selection fitness function, these were calculated to fifty generations; here
we've calculated them to one hundred generations). This lead us to discover a
subtle bias temporal covariance: whether sites that fix or are lost through
time are included or excluded in the temporal covariance calculation can lead
to biases in temporal covariance. We discuss this extensively in the next
section, Supplementary Materials Section \ref{supp:fixation}, but here we note
that all the background selection results exclude fixed and lost sites; these
are marked as missing and excluded from the temporal covariance and $G(t)$
calculations.

@@

\subsubsection{Truncation Selection.}
\gc{We explored how directional truncation selection generated
  temporal covariances.In these simulations we select the top ${\bf X
    \%}$ of the phenotypic distribution of individuals to form the
  next generation. We burnin these simulations {\bf by
    XXXX}. More extreme directional truncation selection generated
  larger initial covariances (Figure
  \ref{suppfig:supp-trunc}B). However, weaker truncation selection
  generated more sustained positive covariances and so had a larger
  long term impact

>>>>

In our analysis of the \textcite{Barghi2019-qy} data, we describe our statistic
$G(t)$ as a lower bound for two reasons: (1) the population is sequenced every
ten generations, meaning the temporal covariances between adjacent generations
cannot contribute to the numerator of $G(t)$ but contribute to the denominator,
and (2) the estimate of $G(t)$ ignores linked selection's contribution to the
per-generation variance in allele frequency change (this is the difference
between our $G(t)$ and $G'(t)$ estimators, the latter of which includes these
variance terms \parencite{Buffalo2019-io}).  To verify that $G(t)$ estimated
every ten generations is indeed a lower bound, we used a simulation procedure
similar to the exponential fitness function simulations (described in
Supplementary Material Section \ref{supp:genetic-arch}), and calculated the
temporal covariances and $G(t)$ both each generation, and every ten
generations. Unlike the simulations described in \ref{supp:genetic-arch}, we
began selection at $10N$ generations, and used trait $V_A = 0.01$ and targeted
$L = 1000$ sites affecting the trait.

@@

We simulated background selection in a 50 megabase region, where deleterious
alleles are randomly introduced by mutation. Following background selection
literature
\parencite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1994-oh,Hudson1995-xc}, we
parameterize the mutation rate as the total number of deleterious mutations
introduced per diploid genome, per generation, and simulate values $U = \{0.5,
1.0, 1.5\}$. We also vary the strength of selection against the deleterious
mutations, $s = \{0.0, 0.01, 0.05, 0.1\}$ (where $s=0$ is a neutral control),
as well as different recombination rates ($r_{bp} = \{10^{-7}, 10^{-8}\}$).
Like other simulations, we burnin the population for $10N$ generations under
backgrounds selection. Overall, we find background selection does create
temporal covariance (Supplementary Material Figure
\ref{suppfig:supp-bgs-covs}), which are stronger under (1) higher deleterious
mutation rates and (2) larger selection coefficients. This latter point
initially seems at odds with background selection theory, as the level of
pairwise diversity in a region under strong background selection is invariant
with respect to the selection coefficient. However, looking at the background
selection $G(t)$ trajectories, we find that over time, the impact of linked
selection created by background selection appears to trend towards an
equilibrium in the $r_{BP} = 10^{-7}$ subfigure, and reaches an equilibrium in
the $r_{BP} = 10^{-8}$ subfigure that seems reasonably invariant to the choice
of $s$ (Supplementary Material Figure \ref{suppfig:supp-bgs-g}). We believe
that these observations can be reconciled by $Cov(\Delta p_t,~ \Delta
p_{t^{\prime}})$ being larger for larger $s$ when $|t^{\prime}-t|$ is small,
but also decaying more rapidly with $|t^{\prime}-t|$, such that the overall
contribution of selection to allele frequency variance is invariant to $s$ (for
strong background selection). However, further future work is needed to fully
explore and understand the temporal covariance dynamics of background
selection.

@@

Additionally, we investigated whether background selection can create
convergence correlations between two replicate populations. Much like the
exponential directional selection and Gaussian stabilizing selection
simulations, we burned in a population for $10N$ generations with background
selection, which continued after the population was split into two replicate
populations. These simulations fixed $U = 1.0$, $r_{BP} = 10^{-8}$, and varied
the replicate population size $n = \{200, 1000\}$. We find that background
selection can create convergence correlations (Supplementary Material Figure
\ref{suppfig:supp-bgs-converg}). We find the convergence correlation is weaker
in smaller replicate population sizes, as there are fewer shared haplotypes
carrying the same deleterious alleles between the two populations.

While investigating} whether different selection coefficients converged to
the same levels of $G(t)$, we extended the number of generations we calculated
our statistics on (e.g. in earlier exponential and Gaussian stabilizing
selection fitness function, these were calculated to fifty generations; here
we've calculated them to one hundred generations). This lead us to discover a
subtle bias temporal covariance: whether sites that fix or are lost through
time are included or excluded in the temporal covariance calculation can lead
to biases in temporal covariance. We discuss this extensively in the next
section, Supplementary Materials Section \ref{supp:fixation}, but here we note
that all the background selection results exclude fixed and lost sites; these
are marked as missing and excluded from the temporal covariance and $G(t)$
calculations.

@@

\subsubsection{Truncation Selection.}

We \vb{also} explored how directional truncation selection generates temporal
covariances. In these simulations we select the top \vb{10\%, 25\%, or 50\%} of
the phenotypic distribution of individuals to form the next generation. We
burnin these simulations \vb{using a neutral burnin routine, where trait
alleles are selectively neutral until directional selection is imposed}. More
extreme directional truncation selection generated larger initial covariances
(Figure \ref{suppfig:supp-trunc}B).  However, weaker truncation selection
generated more sustained positive covariances and so had a larger long term
impact
